
# Creating dashboards {#dashboards}

Dashboards are typically collections of key visualisations and are most effective when they are interactive. Such interactivity is producing by writing R code that is converted into HTML and JavaScript that you can then view in a browser. These dashboards can often have so much functionality that they effectively become web apps. 

## Package options

In order to understand which package to use you need to understand the differences between *static* and *dynamic* webpages. A static webpage loads everything at the start and never changes; this loading could include interactive content such as D3 visualisations or leaflet maps which are all run client-side (the processing is done in the browser). For a dynamic webpage the content can be reloaded; for example, data could be pulled from a database or items changed due to the running of R code, normally on a remote server. Dynamic webpages are more flexible and allow you to work with larger datasets but require a web server, whereas static web pages can be run independently, not requiring the user to have R themselves or the web page to be hosted on a server.

### Flexdashboard

The `flexdashboard` package in R extends R markdown (used to create reports in R) to include dashboard layouts. In effect these layouts consists of elements (be they tables, `ggplot2` charts, `plotly` charts, or `leaflet` maps) which are arranged in rows and columns.

When using `flexdashboard` on its own these separate elements cannot talk to one another, but the resulting dashboard is a self-contained HTML file that can be emailed to a customer.

Examples can be found on the [flexdashboard website](https://rmarkdown.rstudio.com/flexdashboard/). These dashboard layouts can be combined with `crosstalk` and `shiny` for additional functionality.

### Crosstalk

The `crosstalk` package allows HTML content produced by certain packages to interact. This means that if you click on one element (which could be a slider to select data ranges), the others will update. At the time of writing `crosstalk` only works with `plotly`, `leaflet`, `DT`, `SummaryWidget`, and `rgl`.

Combining `crosstalk` with `flexdashboard` allows the creation of dashboards where the elements of a dashboard can simultaneously update based on user interaction with one particular element, and the output produced is still an easily sharable HTML file.

More information can be found on the [crosstalk website](https://rstudio.github.io/crosstalk/widgets.html).

### Shiny

The `shiny` package allows the creation of very powerful dashboards where user interaction (for example, clicking a button) can result in arbitary R code being run behind the scenes and the dashboard being updated accordingly. This allows almost limitless functionality within the dashboard; models could be rerun, new data called from a database *etc*. However, `Shiny` requires R to be running somewhere, ideally on a server but it could be on a local computer.

At the moment CoDE does not have a web server on which to run these dashboards. The ways in which you could currently share a `Shiny` dashboard are:

- Get your customer to install RStudio on their computer and run the dashboard code themselves.
- Find an unused networked computer to host the dashboard on.
- Pay for an internal server to run the application on.

## Getting started with Shiny

Shiny apps have two main components, the *user interface* (file saved as ui.R) and the *reactive server* (file saved as server.R).

### ui.R

The user interface script holds all information on the static (permanent) design and layout of the app.

There are a variety of different packages and pre-set designs available to use - to name a few:

- navBarPage - *good for tabs*
- fluidPage - *good when using on different device types*
- dashboardpage - *general purpose*

Each of these designs allows you to easily create a sidebar/mainPanel layout for easy dashboard use.

This is where you tell Shiny where to position all visible elements:

- An *input* is a reference for `Shiny` to take the value chosen by the user to change another element on the page (usually added on the ui.R side)
- An *output* is a widget, plot, table, map, chart or input box (tip of the iceberg!) that can change depending on any other input on the page
- The server script (see below) handles all *interactive elements* (inputs and outputs), updating them on user changes (AKA reactivity)

*Note: It is possible to make pretty much any ui element reactive by adding an Output function (with an id) on the ui side, and then building it out on the server side.*

```{r ui-script, eval=FALSE}

library(shiny)

# Define UI for app that draws a histogram ----
ui <- fluidPage(

  # App title ----
  titlePanel("Hello Shiny!"),

  # Sidebar layout with input and output definitions ----
  sidebarLayout(

    # Sidebar panel for inputs ----
    sidebarPanel(

      # Input: Slider for the number of bins ----
      sliderInput(inputId = "bins",
                  label = "Number of bins:",
                  min = 1,
                  max = 50,
                  value = 30)

    ),

    # Main panel for displaying outputs ----
    mainPanel(

      # Output: Histogram ----
      plotOutput(outputId = "distPlot")

    )
  )
)

```

### server.R

As mentioned above, the server element of the Shiny app exists to make it *interactive*.

In order to reference *clicks/choices/hovers/selections* from other elements in the app, use *input\$id* (e.g. "input\$bins" as  below).

You can prevent elements from updating each time or make them wait for dependent inputs before rendering by using the `isolate()` and `req()` functions.

```{r server-script, eval=FALSE}

# Define server logic required to draw a histogram ----
server <- function(input, output) {

  # Histogram of the Old Faithful Geyser Data ----
  # with requested number of bins
  # This expression that generates a histogram is wrapped in a call
  # to renderPlot to indicate that:
  #
  # 1. It is "reactive" and therefore should be automatically
  #    re-executed when inputs (input$bins) change
  # 2. Its output type is a plot
  output$distPlot <- renderPlot({

    x    <- faithful$waiting
    bins <- seq(min(x), max(x), length.out = input$bins + 1)

    hist(x, breaks = bins, col = "#75AADB", border = "white",
         xlab = "Waiting time to next eruption (in mins)",
         main = "Histogram of waiting times")

    })

}


```

### Running the app

It's good practice to separate your ui and server scripts into separate files and then use a `runApp()` function (only requires the folder location) to draw them together.

However, for immediate learning processes here the full script that you need to run the app is shown above.

```{r full-script, eval=FALSE}

library(shiny)

# Define UI for app that draws a histogram ----
ui <- fluidPage(

  # App title ----
  titlePanel("Hello Shiny!"),

  # Sidebar layout with input and output definitions ----
  sidebarLayout(

    # Sidebar panel for inputs ----
    sidebarPanel(

      # Input: Slider for the number of bins ----
      sliderInput(inputId = "bins",
                  label = "Number of bins:",
                  min = 1,
                  max = 50,
                  value = 30)

    ),

    # Main panel for displaying outputs ----
    mainPanel(

      # Output: Histogram ----
      plotOutput(outputId = "distPlot")

    )
  )
)

# Define server logic required to draw a histogram ----
server <- function(input, output) {

  # Histogram of the Old Faithful Geyser Data ----
  # with requested number of bins
  # This expression that generates a histogram is wrapped in a call
  # to renderPlot to indicate that:
  #
  # 1. It is "reactive" and therefore should be automatically
  #    re-executed when inputs (input$bins) change
  # 2. Its output type is a plot
  output$distPlot <- renderPlot({

    x    <- faithful$waiting
    bins <- seq(min(x), max(x), length.out = input$bins + 1)

    hist(x, breaks = bins, col = "#75AADB", border = "white",
         xlab = "Waiting time to next eruption (in mins)",
         main = "Histogram of waiting times")

    })

}

shinyApp(ui = ui, server = server)

```


## Advanced considerations

Shiny's capabilities are effectively endless; here are some things to consider when designing your dashboards.

### User Experience:  
  
- [Linking Plotly charts as an input for other elements](https://plot.ly/r/shiny-coupled-events/)  
- [Choosing icons](https://fontawesome.com/icons?from=io)  
- [Running javascript code with shinyJS](https://github.com/daattali/shinyjs)  
- [Improving the user experience with progress indicators](https://shiny.rstudio.com/articles/progress.html)  
- [Using downloadHandler to download tables](https://shiny.rstudio.com/articles/download.html)  
- [Downloading ggplots with ggsave](https://stackoverflow.com/questions/14810409/save-plots-made-in-a-shiny-app)  
- [Showing elements based on conditions](https://shiny.rstudio.com/gallery/conditionalpanel-demo.html)  


### Speed:

- Run all the aggregations, data processing and queries in advance, then run `Shiny` from condensed datasets at an individual level.
- Add the data used into a database, then query as and when needed instead of running in memory.
- Use the `req()` function inside `render({})` functions on your server.R script to stop the app trying to draw plots/tables until the required reactive data/filters have already been sorted out.
- Use the `isolate()` function to stop input changes affecting every output they're linked to every time.


### Other considerations:  
  
- [Separating code for complex dashboards with Shiny Modules](https://shiny.rstudio.com/articles/modules.html)  
- [Saving the state of a Shiny application with Bookmarking](https://shiny.rstudio.com/articles/bookmarking-state.html)  
- [Tracking client usage with session$clientData](https://shiny.rstudio.com/articles/client-data.html)  
- [Linking to Google Analytics for broader tracking](https://shiny.rstudio.com/articles/google-analytics.html)  
- [Debugging Shiny applications](https://shiny.rstudio.com/articles/debugging.html)  
- [Server-to-client custom messages](https://shiny.rstudio.com/gallery/server-to-client-custom-messages.html)  
- [Deploying Shiny apps on the web (with RStudio's own services)](https://shiny.rstudio.com/articles/deployment-web.html)  


## Learning resources
  
- [**Shiny Official Site**](https://fontawesome.com/icons?from=io)  
- [**Shiny Cheat Sheet**](https://shiny.rstudio.com/articles/cheatsheet.html)  
  
**Tutorials**  
  
- [Shiny Official Tutorial](https://shiny.rstudio.com/tutorial/written-tutorial/lesson1/)  
- [Datacamp Tutorial](https://www.datacamp.com/courses/building-web-applications-in-r-with-shiny)  
- [Mango Solutions Tutorial](https://github.com/MangoTheCat/shiny_beyond_the_basics/blob/master/Workshop%202%20-%20Shiny%20-%20Beyond%20the%20Basics.pdf)  

**Visualisation packages**

- [**DT** (best for interactive tables!)](https://rstudio.github.io/DT/shiny.html)  
- [**Plotly** (use ggplotly() function to convert ggplot to interactive charts)](https://plot.ly/r/shiny-tutorial/)  
- [**Leaflet** (maps)](https://rstudio.github.io/leaflet/shiny.html)  
- [**d3**](http://www.htmlwidgets.org/showcase_networkD3.html)  
- [**Google Charts**](https://github.com/mages/googleVis/)
  
**Inspiration**  
  
- [RStudio App Gallery](https://shiny.rstudio.com/gallery/)  
- [Shiny Extended Gallery](https://shiny.rstudio.com/gallery/see-more.html)  
- [Show Me Shiny Gallery](https://www.showmeshiny.com/)  
  
**Influencers**  
  
- [Dean Attali](https://deanattali.com/blog/advanced-shiny-tips/)  
- [Joe Cheng](https://twitter.com/jcheng?lang=en)  
  
**HMRC Examples**  
  
- [shinyTemplate](https://dataexploitation.shinyapps.io/shinytemplate/) 

