
# Working with dates

Dates and date-times are a base type in R and can be used in arithmetic and within logical expressions.

If you import date using one of the functions within `readr` it will try to infer the date format; turning the date field into a date variable where possible. Otherwise the field would appear as a string and you would need to tell R that this string is a date by specifying the format. For example:
```{r eval=FALSE}
string_date <- "30-01-2010"
date <- as.Date(string_date, format = "%d-%m-%Y")
```

Similarly date-times can be read in as follows:
```{r eval=FALSE}
string_datetime <- "30-01-2010 10:30:05"
date_time <- as.POSIXct(string_datetime, format = "%d-%m-%Y %H:%M:%S")
```

The `lubridate` package contains a variety of ultities that make it easier to work with dates and times. For example, it offers a series of functions that infer the date for you:
```{r eval=FALSE}
library(lubridate)

date_1 <- ymd("2010 April 01")
date_2 <- mdy("Apr 01, 10")
date_3 <- dmy("1st April 2010")
```
and functions to work with intervals between date ranges, and to easily add a set number of days, weeks or months to dates.

Resources:

* RStudio cheat sheet on working with dates and times [(download here)](https://github.com/rstudio/cheatsheets/raw/master/lubridate.pdf)
* Quick-R guide to [date values](https://www.statmethods.net/input/dates.html)
