
# Programming


R is a fully fledged programming language. This section is a very quick introduction to how common programming constructs are implemented within R.

### Defining your own function

To define a function called `function_name` you would use the following syntax.
```{r eval=FALSE}

function_name <- function(arguement_1, arguement_2){
  ...
  return(result)
}

```

The final return statement is optional; if it is not present the function will return the value of the last expression evaluated. These functions can then be called in the usual way, *e.g.* `function_name(arg1, arg2)`.

### If statements

The syntax for *if statements* is the following:
```{r eval=FALSE}

if (condition_1) {
  ...
} else if (condition_2) {
  ...
} else {
  ...
}


```
where the conditions are any logical tests.

### For loops

R does give you the option to use *for loops* but also provides a number of other ways to iterate through a series of values. Consider the following example where we first define a vector of strings.
```{r eval=FALSE}

vec_a <- c("string", "str", "s")

```

We want to find the length of each of these strings. You could use a for loop to iterate over all the elements in this vector.
```{r eval=FALSE}

for (i in vec_a){
  print(nchar(i))
}


```
(Another helpful syntax is that, for example, `1:10` produces a list of the numbers between one and ten that can be iterated through.)

However, many functions in R have been designed to work directly with vectors and produce vectors as outputs. `nchar()` is one such function.

```{r eval=FALSE}
nchar(vec_a)
```

Let's imagine that `nchar` did not work on a vector, that it would only take a single value as input. If we wanted to iterate over our vector and for each element apply the `nchar` function we could use the functions `lapply`, `sapply`, and `vapply`. 
```{r eval=FALSE}

# The lapply function produces a list
lapply(vec_a, nchar)

# The sapply function tries to simplify to a vector
sapply(vec_a, nchar)

# It is recommended to use vapply where you specify the output type
vapply(vec_a, nchar, integer(1))

```

The package `purrr` (part of the `tidyverse`) provides newer, more consistent replacements for the `apply` family of functions. These are the `map` functions.
```{r eval=FALSE}
library(purrr)

# map produces a list
map(vec_a, nchar)

# map_int produces a vector of integers
# map_dble, map_char and map_lgl exist for other data types
map_int(vec_a, nchar)

```


Resources:

* *R for Data Science* has a good [chapter on programming](http://r4ds.had.co.nz/program-intro.html) which introduces many important concepts.
* [Programming with R](https://swcarpentry.github.io/r-novice-inflammation/) has a number of episodes demonstrating different aspects of R programming.
* RStudio cheat sheet on applying functions with `purrr` [(Download here)](https://github.com/rstudio/cheatsheets/raw/master/purrr.pdf).
* More advanced topics such as different programming styles and writing more performant code are discussed in [Advanced R](http://adv-r.had.co.nz/).
* Creating your own R package is [discussed here](http://r-pkgs.had.co.nz/).
* Debugging code is discussed [in these articles](https://support.rstudio.com/hc/en-us/articles/205612627-Debugging-with-Rstudio).
* The [tidyverse style guide](http://style.tidyverse.org/) is a commonly followed style guide.


