
# Taking samples

This section focuses on generating samples from data. Most types of sampling can be done by manipulating the data using functions from the packages `dplyr` or `purrr`. These should be enough for most purposes. 

More advanced sampling techniques can be found in the `sampling` package or in the `boot` package (which focuses on bootstrapping methods). If you already have survey data, the `survey` package allows you to calculate statistics taking into account factors such as different finite population corrections.

Examples of sampling the `df` dataframe:

### Simple random sampling

```{r eval=FALSE}
library(dplyr)

my_sample <- 
  df %>%
	sample_n(100, replace = TRUE)
```

###Stratified sampling (weighted)

```{r eval=FALSE}
library(dplyr)

my_sample <- 
  df %>%
  group_by(group_col) %>%
  mutate(num = n()) %>%
	sample_n(100, weight = num)
```

### Cluster sampling

```{r eval=FALSE}
library(dplyr)

# Weights for each class can also be included
class_names <- c("class1", "class2", "class3", "class4")
class_sample <- sample(class_names, 3)

my_sample <- 
  df %>%
  filter(class_col %in% class_sample)
```

### Systematic sampling

```{r eval=FALSE}
library(purrr)

n <- 100
every <- 10
rows <- nrow(df)
start <- sample(1:rows, size = 1)

# Generate indices for the sample
indx <- seq(from = start, to = (start + (n-1)*every), by = every)

# If index larger than the number of rows, cycle round the dataframe
indx2 <-
  indx %>%
  map_dbl(~ (.x -1) %% rows + 1 )

# Subset original data
my_sample <- df[indx2, ]
```

Note that the `sample_n` command might not work with database connections.


Resources:

* `dplyr` [documentation](https://dplyr.tidyverse.org/reference/sample.html) on sampling row
* Quick-R [guide to bootstrapping](https://www.statmethods.net/advstats/bootstrapping.html) 
* `survey` package [documentation](https://dplyr.tidyverse.org/reference/sample.html)


